---
title: "HW03 - Ranking NBA Teams"
author: "Bryant Luong"
date: "10/13/2017"
output: github_document
---
## Setup
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::read_chunk(path = '../code/make-teams-table.R')
```
```{r packages}
```

## Data Preparation
```{r tables}
# create 'NBAroster' and 'NBAstats' data frames from make-teams-table.R script
```
```{r teams}
# create 'teams' data frame
```

## Basic Ranking: By Salary
```{r rbs}
# rank team by salary 
teamsBySalary.df <- arrange(teams, salary)

# make 'teamBySalary.df$team' an ordered factor
teamsBySalary.df$team <- factor(teamsBySalary.df$team, levels = teamsBySalary.df$team)

# create horizontal barchart with vertical red line representing average team salary
p <- ggplot(data = teamsBySalary.df,
            mapping = aes(team, salary))
       
pbl <- p + geom_bar(stat = 'identity', 
                    fill = "grey") + coord_flip() + labs(x = 'Team',
                                                         y = 'Salary (in millions)',
                                                         title = 'NBA Teams ranked by Total Salary')

pbl + geom_hline(aes(yintercept=mean(teamsBySalary.df$salary)), 
                 color = rgb(1, 0, 0, 0.5), 
                 lty = 1, lwd = 1)
```

### Provide a concise description of this basic ranking.

The mean team salary was approximately $91 million and 6 teams were very close to the mean. 
The standard deviation was $15.76 million.

The Cleveland Cavaliers had the largest team salary of $125.79 million, which is 2 SDs above
the mean. The Philadelphia 76ers had the smallest team salary of $55.78 million, which is 2 SDs
below the mean.

The team salaries change the fastest around \$60 to \$80 million. After \$80 million a year, 
the increase in team salary between teams is smaller. However, the biggest difference between 
2 neighboring teams on the barplot is between the highest and second-highest team salaries. 
The Cavaliers spend almost $40 million more than the Clippers.

## Basic Ranking: By Points
```{r rbp}
# rank team by total points
teamsByPoints.df <- arrange(teams, points)

# make 'teamByPoints.df$team' an ordered factor
teamsByPoints.df$team <- factor(teamsByPoints.df$team, levels = teamsByPoints.df$team)

# create horizontal barchart with vertical red line representing average team total points
p <- ggplot(data = teamsByPoints.df,
            mapping = aes(team, points))

pbl <- p + geom_bar(stat = 'identity', 
                    fill = "grey") + coord_flip() + labs(x = 'Team', 
                                                         y = 'Total Points',
                                                         title = 'NBA Teams ranked by Total Points')


pbl + geom_hline(aes(yintercept=mean(teamsByPoints.df$points)), 
                 color = rgb(1, 0, 0, 0.5), 
                 lty = 1, lwd = 1)
```

### Provide a concise description of this basic ranking.

In the 2017 season, the Sacramento Kings scored the least points and the Golden State Warriors scored the most points.
The Warriors scored 9,473 points and the Kings scored only 6,348, so the Warriors scored almost 50% more points. 
The Warriors even scored 562 more points than the Clippers, which is the biggest difference between 2 neighboring teams
on the barplot.

The mean number of points was 8,035 and the standard deviation was 710 points. The Warriors are 2 SDs above the mean.

The Clippers is the only team that had the same ranking by salary and by points. By these two metrics, they're 2nd in the NBA.

## Basic Ranking: By Efficiency
```{r rbe}
# rank team by efficiency
teamsByEff.df <- arrange(teams, efficiency)

# make 'teamByEff.df$team' an ordered factor
teamsByEff.df$team <- factor(teamsByEff.df$team, 
                             levels = teamsByEff.df$team)

# create horizontal barchart with vertical red line representing average team total points
p <- ggplot(data = teamsByEff.df,
            mapping = aes(team, efficiency))

pbl <- p + geom_bar(stat = 'identity', 
                    fill = "grey") + coord_flip() + labs(x = 'Team',
                                                         y = 'Efficiency',
                                                         title = 'NBA Teams ranked by Efficiency')

pbl + geom_hline(aes(yintercept=mean(teamsByEff.df$efficiency)), 
                 color = rgb(1, 0, 0, 0.5), 
                 lty = 1, lwd = 1)
```

### Provide concise descriptions of the basic rankings.

Comparing the rankings by salary and by efficiency, 2 teams have the same ranking, the Lakers and Cavaliers. 
The Lakers are ranked #20 and the Cavaliers are ranked #1. 

Comparing the rankings by points and by efficiency, 0 teams have the same ranking.

The mean efficiency is 149 and the standard deviation is 11.28. The Cavaliers have the maximum efficiency of 
177.86 and the Magics have the minimum efficiency of 125.14. 22 teams are within 1 SD and 5 teams 
are above 1 SD.

## Principal Components Analysis
```{r pca}
# create data frame for PCA
pca.df <- teams %>% select(team, 
                           points3, 
                           points2, 
                           free_throws, 
                           off_rebounds, 
                           def_rebounds, 
                           assists, 
                           steals, 
                           blocks, 
                           turnovers, 
                           fouls)

# compute PCA
pcs <- prcomp(pca.df[-1], scale. = TRUE)

# create data frames for eigenvalues, their proportions, and cumulative proportions
eigs <- round(pcs$sdev^2, digits = 4)
prop <- round(pcs$sdev^2 / sum(pcs$sdev^2), digits = 4)
cumprop <- cumsum(prop)

# create data frame with all 3 metrics
eigs.df <- data.frame(eigs, prop, cumprop)
eigs.df

# create data frame that contains only PC1 and PC2 values of each team
pc1pc2 <- data.frame('PC1' = pcs$x[,1], 
                     'PC2' = pcs$x[,2],
                     'team' = teams$team)

# create scatterplot of teams using PC1 and PC2
ggplot(data = pc1pc2, 
       aes(x = PC1, y = PC2, label = team)) + geom_text() + geom_hline(yintercept = 0, color = 'grey') + geom_vline(xintercept = 0, color = 'grey') + labs(title = 'PCA plot (PC1 and PC2)')
```

### Interpret the scatterplot. 

The purpose of PCA is to find the basis that captures the most variation in the data and to reduce the dimensions of data.
PCA will guide us to the teams that are most different from other teams by giving us principal components (PC).
Principal components can be understood as summaries of all dimensions of the data.
By plotting PC1 and PC2 together, we are able to **see** the performances of teams because our data has the ten most critical
metrics of game play: 3-pointers, 2-pointers, free throws, rebounds, assists, steals, turnovers, blocks, and fouls.

The teams that are inside -3 and 3 on the PC1 axis have performed similarly. The unusual teams are GSW, NOP, DAL, and SAC.
To further understand why these teams are outliers, we first look at the weights for PC1 to understand how each game play
statistic contributes to the PC score. We know the PC scores are a linear combination of the 10 game play metrics. 
The weights for PC1 are:

variable     | weight 
------------ | ------
points3      | 0.1122
points2      | 0.3602
free_throws  | 0.3228
off_rebounds | 0.3029
def_rebounds | 0.3719
assists      | 0.3125
steals       | 0.3447
blocks       | 0.3162
turnovers    | 0.3354
fouls        | 0.3073

The game play statistics that contribute most to the PC score are defensive rebounds, 2-pointers, and turnovers. Let's
check these statistics for GSW, NOP, DAL, and SAC.
```{r def}
pca.df %>% group_by(team) %>% arrange(def_rebounds)
pca.df %>% group_by(team) %>% arrange(desc(def_rebounds))
```
The data above tells us that SAC, DAL, and NOP are the 3 teams with the least defensive rebounds and GSW had the most defensive rebounds. Let's
do the same search for 2-pointers and turnovers.
```{r 2p}
pca.df %>% group_by(team) %>% arrange(points2)
pca.df %>% group_by(team) %>% arrange(desc(points2))
```
The data above tells us that SAC, DAL, and NOP are among the teams that scored the fewest 2-pointers and GSW is in the top 5. 
```{r to}
pca.df %>% group_by(team) %>% arrange(turnovers)
pca.df %>% group_by(team) %>% arrange(desc(turnovers))
```
The data above tells us that DAL, NOP, and SAC are in the top 5 for fewest turnovers while GSW is in the top 3 for the most turnovers. 

We can do a similar analysis for PC2 for teams greater than 2 and less than -2. These teams are PHO, HOU, and GSW.

## Index based on PC1

```{r scaled_pc1}
# create index based on PC1
s1 <- data.frame('s1' = 100 * (pcs$x[,1] - min(pcs$x[,1])) / (max(pcs$x[,1]) - min(pcs$x[,1])),
                 'team' = teams$team)

# rank team by scaled PC1
s1 <- arrange(s1, desc(s1))

# make 'teams' an ordered factor
s1$team <- factor(s1$team, levels = s1$team)

# create horizontal bar chart of scaled PC1 to visualize ranking  
p <- ggplot(data = s1,
            mapping = aes(team, s1))
p + geom_bar(stat = 'identity', 
             fill = "grey") + coord_flip() + labs(x = 'Team',
                                                  y = 'First PC (scaled from 0 to 100)',
                                                  title = 'NBA Teams ranked by scaled PC1')
```

### Provide a brief description of the PC1 index to rank the teams.

The PC1 index uses the first principal component to create a 0-100 scale and each team is ranked relative to the minimum 
first principal component. The width of the scale is determined by the minimum and maximum principal component score of the teams.
Because the first principal component score of each team is calculated using the same set of weights, the variation we see
is from the game play statistics. As we can see in the table above, the principal component weights are all between 0.3 and 0.4 
so this gives us confidence that there isn't too much bias towards 1 game play statistics. In fact, the GSW have the best 3-point
statistics but that statistic only gets weighted by 0.11. With this mind, they still come up at the top!

## Comments and Reflections

**1. Was this your first time working on a project with such file structure? If yes, how do you feel about it?**

It was my first time and it felt very organized. It made this mini project easier to manage.

**2. Was this your first time using relative paths? If yes, can you tell why they are important for reproducibility purposes?**

It was my fist time using relative paths and I now understand its usefulness. Relative paths ensure files in the scripts can be located and used.

**3. Was this your first time using R script?**

No, we used it in lab!

**4. What was hard, even though you saw them in class/lab?**

There are ggplot functions that are not intuitive to me. 

**5. What was easy?**

The first parts of the mini project were easy. 

**6. Did anyone help you complete this assignment?**

No.

**7. How much time did it take to complete this HW?**

8 hours.

**8. What was the most time consuming part?**

Figuring out why my barplot was being automatically sorted alphabetically. 

**9. Was there anything interesting?**

PCA is very cool and useful.

