---
title: "HW 02 - Basics of Data Frames"
author: "Bryant Luong"
output:
  html_document:
    theme: lumen
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    number_sections: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
# add libraries
library(readr)
library(dplyr)
```
# Create a data dictionary  

The data dictionary is stored at the following link:

https://github.com/earnestpointers/stat133-hws-fall17/blob/master/hw02/data/nba2017-player-statistics-dictionary.md

# Importing the data into R
```{r read}

# Use read.csv() to read csv into R and check with str()
base <- read.csv(file = "nba2017-player-statistics.csv",
                 colClasses = c('character', 
                                'character', 
                                'factor',
                                'character', 
                                'double', 
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer',
                                'integer'))
str(base)

# Use read_csv() to read csv into R and check with str()
rdr <- read_csv(file = "nba2017-player-statistics.csv", 
                col_types = cols( Player = col_character(), 
                                  Team = col_character(), 
                                  Position = col_factor(levels = c('C',
                                                                   'PF',
                                                                   'PG',
                                                                   'SF',
                                                                   'SG')),
                                  Experience = col_character(), 
                                  Salary = col_double(), 
                                  Rank = col_integer(),
                                  Age = col_integer(),
                                  GP = col_integer(),
                                  GS = col_integer(),
                                  MIN = col_integer(),
                                  FGM = col_integer(),
                                  FGA = col_integer(),
                                  Points3 = col_integer(),
                                  Points3_atts = col_integer(),
                                  Points2 = col_integer(),
                                  Points2_atts = col_integer(),
                                  FTM = col_integer(),
                                  FTA = col_integer(),
                                  OREB = col_integer(),
                                  DREB = col_integer(),
                                  AST = col_integer(),
                                  STL = col_integer(),
                                  BLK = col_integer(),
                                  TO = col_integer()))
str(rdr)
```

# Right after importing the data
```{r processing}
# Replace occurences of "R" with 0 
base$Experience[base$Experience == "R"] = 0

# Convert Experience column from character to integer type
base$Experience <- as.integer(base$Experience)
str(base)
```

# Performance of players
```{r eff}
# Add Missed_FG to base
base$Missed_FG <- (base$FGA - base$FGM)

# Add Missed_FT to base 
base$Missed_FT <- (base$FTA - base$FTM)

# Add PTS to base
base$PTS <- (3*base$Points3 + 2*base$Points2 + base$FTM)

# Add REB to base
base$REB <- (base$OREB + base$DREB)

# Add MPG to base
base$MPG <- (base$MIN - base$GP)

# Add EFF to base
base$EFF <- c(base$PTS + 
              base$REB +
              base$AST +
              base$STL +
              base$BLK -
              base$Missed_FG -
              base$Missed_FT -
              base$TO) / base$GP

# Summary statistics of EFF
summary(base$EFF)

# Histogram of EFF
hist(x = base$EFF,
     ylim = c(0, 200),
     xlab = 'EFF',
     ylab = 'Frequency',
     main = 'Histogram of Efficiency (EFF)',
     col = rgb(0.32, 0.32, 0.32, 0.50))

# Top 10 players by EFF
top.ten <- arrange(.data = base, desc(EFF)) %>% slice(1:10) %>% select(Player, Team, Salary, EFF)
top.ten

# Players with negative EFF
neg.eff <- arrange(.data = base) %>% filter(EFF < 0) %>% select(Player)
neg.eff

# Compute correlation coefficient between EFF and all the variables in EFF formula
cor.pts <- cor(x = base$EFF,
               y = base$PTS)

cor.reb <- cor(x = base$EFF,
               y = base$REB)

cor.stl <- cor(x = base$EFF,
               y = base$STL)

cor.ast <- cor(x = base$EFF,
               y = base$AST)

cor.blk <- cor(x = base$EFF,
               y = base$BLK)

cor.missed_fg <- cor(x = base$EFF,
                     y = base$Missed_FG)

cor.missed_ft <- cor(x = base$EFF,
                     y = base$Missed_FT)

cor.to <- cor(x = base$EFF,
                     y = base$TO)

# Create data frame with all coefficients
all.cor <- data.frame(c(cor.pts, 
                        cor.reb, 
                        cor.stl, 
                        cor.ast, 
                        cor.blk, 
                        cor.missed_fg * -1, 
                        cor.missed_ft * -1, 
                        cor.to * -1))

# Name column 'coefficients'
attributes(all.cor)$names[1] <- 'coefficients'

# Sort coefficients by decreasing value
sorted.all.cor <- arrange(.data = all.cor,
                          desc(coefficients))

# Display coefficients
sorted.all.cor

# Create vector with variable names for x-axis of plot           
cnames = c('PTS', 
           'REB', 
           'STL', 
           'AST', 
           'BLK', 
           'Missed_FT', 
           'Missed_FG', 
           'TO')

# Create barplot of efficiency and salary
barplot(height = as.vector(sorted.all.cor[,1]),
        main = "Correlations between Player Stats and EFF",
        names.arg = cnames,
        axis.lty = 1,
        ylim = c(-1, 1),
        cex.names = 0.55,
        cex.axis = 1,
        border = 'NA',
        col = c('grey',
                'grey',
                'grey',
                'grey',
                'grey',
                'red',
                'red',
                'red'))
```

# Efficiency and Salary

## All Players 
```{r scatterplot}
# Add scatterplot of efficiency and salary
plot(x = base$EFF,
     y = base$Salary/1000000,
     xlab = "Efficiency",
     ylab = "Salary ($, millions)",
     main = 'Scatterplot of Efficiency and Salary with Lowess line for All Players',
     pch = 16,
     col = rgb(0.8, 0.21, 0.3, 0.4))

# Add lowess line to plot
lines(lowess(x = base$EFF, y = base$Salary/1000000),
      lty = 1,
      lwd = 3,
      col = '#458b00')

# Add 'lowess' label to plot
text(x = 33, 
     y = 30, 
     labels = 'lowess', 
     col = '#458b00', 
     cex = 0.8)

# Compute correlation coefficient between efficiency and salary
cor.eff.sal <- cor(x = base$EFF,
                   y = base$Salary)
cor.eff.sal
```

### What can you say about the relationship between these two variables?

A correlation coefficient of 0.65 suggests the relationship between efficiency and salary is weak. 
In other words, you'd likely predict the wrong salary if you're given an efficiency value. 
It seems salary is dependent on other variables that aren't included in the efficiency equation.

## Players with More Experience
```{r pros}
players2 <- filter(base, MPG >= 20)

# Add scatterplot of efficiency and salary
plot(x =players2$EFF,
     y =players2$Salary/1000000,
     xlab = "Efficiency",
     ylab = "Salary ($, millions)",
     main = 'Scatterplot of Efficiency and Salary with Lowess line for Established Players',
     pch = 16,
     col = rgb(0.21, 0.8, 0.3, 0.4),
     cex.main = 0.95)

# Add lowess line to plot
lines(lowess(x = players2$EFF, 
             y = players2$Salary/1000000),
      lty = 1,
      lwd = 3,
      col = '#8b0046')

# Add 'lowess' label to plot
text(x = 33, 
     y = 30, 
     labels = 'lowess', 
     col = '#8b0046', 
     cex = 0.8)

# Compute correlation coefficient between efficiency and salary
cor.eff.sal <- cor(x = players2$EFF,
                   y = players2$Salary)
cor.eff.sal
```
### What can you say about these two variables for the established players?

The correlation coefficient of efficiency and salary for established players is 0.63. 
This coefficient is 0.002 lower than the coefficient for all players. 
The correlation is even weaker for established players. 
Salary may depend on other variables that aren't included in the efficiency equation.
All the points that aren't close to the lowess line tell us that there is variation on how much players earn per efficiency value.
There is a lot of variation in salary for players who have efficiency values between 5 and 14. 
Players who have efficiency values above 14 earn more per efficiency value on average. 


# Comments and Reflection

**1. What was hard?**

There is a subtlety to data frames that I sometimes trip over. This homework helped find that subtlety.

**2. What was easy?**

The plots were easy because we learned them on the last homework.

**3. Did you need help?**

Yes. The homework instructions aren't clear when it tells us to add columns for Missed_FG, Missed_FT, MPG, etc. It should be explicit that we are supposed to calculate them with the data in the table.

**4. How much time did it take?**

4 hours.

**5. What was the most time consuming?**

Formatting the code.

**6. Was there anything that you did not understand?**

I had to think about why 3 correlation coefficients are supposed to be negative.

**7. Was there anything frustrating?**

No.

**8. Was there anything exciting?**

Yes. We are using larger data sets!

